---
title: "Meta-analysis: Immune-checkpoint blockade (ICB) data"
author: "Farnoosh Babazadeh, Minoru Nakano, Sisira Kadambat Nair, and Benjamin Haibe-Kains "
date: ''
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(meta)
library(metafor)
library(ggplot2)
library(survival)
library(survminer)
library(readr)
library(magrittr)
library(forestplot)
library(stringr)

```

This RMarkdown script shows Meta-analysis of subset of ICB datasets. The script is organized as follows:

1. Load ICB datasets
2. Logistic regression analysis
3. Meta-analysis
 
## Load data

To download and extract TSV version of ICB data from Zenodo (https://zenodo.org/record/7058399/files/), follow the script named download_tsv_data.R file at "~/Code/Download_Data".  

We read RNA-seq expression, clinicopathological characteristics, and gene meta data for the following selected ICB studies to include in meta-analysis. 

1. Braun (PMID 32472114)
2. Gide (PMID 30753825)
3. Hugo (PMID 26997480)
4. Kim (PMID 30013197)
5. Miao1 (PMID 29301960)
6. Nathanson (PMID 27956380) 
7. Riaz (PMID 29301960)
8. Snyder (PMID 28552987) 
9. Van Allen (PMID 26359337)
10. VanDenEnde (PMID 33504550)

As an example, we share the codes to get read the RNA-seq expression, clinicopathological characteristics, and gene meta data usign ICB Braun data. The Braun study includes 319 patients with clinical data and 40,994 genes. While only 181 patients have both clinical data and RNA-seq expression data. 

Note that the expression data is $log2(\text{TPM} + 0.001)$, where TPM represents transcripts per million. 


### Braun study


```{r load ICB data Braun}

app_dir <- str_split(rstudioapi::getActiveDocumentContext()$path, 'icb_data_exploration_and_analysis')[[1]][1]
dir <- file.path(app_dir, 'icb_data_exploration_and_analysis', '.local_data')

# clinicopathological characteristics  
dat_sample_braun = read.table(file.path(dir, 'ICB_Braun/ICB_Braun_metadata.tsv'), sep="\t") %>% as.data.frame()
dim(dat_sample_braun)


# gene meta data
dat_gene_braun = read.table(file.path(dir, 'ICB_Braun/ICB_Braun_expr_genes.tsv'), sep="\t") %>% as.data.frame()
dim(dat_gene_braun)

# RNA-seq expression data (i.e., log2(TPM + 0.001)) 
dat_exp_braun = read.table(file.path(dir, 'ICB_Braun/ICB_Braun_expr.tsv'), sep="\t") %>% as.data.frame()
dim(dat_exp_braun)

# Get the subset of data with both clinical and expression data.

dat_sample_braun = dat_sample_braun[rownames(dat_sample_braun) %in% colnames(dat_exp_braun), ]

dim(dat_sample_braun)

dat_sample_braun <- dat_sample_braun[order(rownames(dat_sample_braun)), ]
dat_exp_braun <- dat_exp_braun[, order(colnames(dat_exp_braun))]

```

```{r load ICB data, include=FALSE}

##################################
### Nathanson study
##################################

# clinicopathological characteristics  
dat_sample_nathanson = read.table(file.path(dir, 'ICB_Nathanson/ICB_Nathanson_metadata.tsv'), sep="\t") %>% as.data.frame()
dim(dat_sample_nathanson)

# gene meta data
dat_gene_nathanson = read.table(file.path(dir, 'ICB_Nathanson/ICB_Nathanson_expr_genes.tsv'), sep="\t") %>% as.data.frame()
dim(dat_gene_nathanson)

# RNA-seq expression data (i.e., log2(TPM + 0.001)) 
dat_exp_nathanson = read.table(file.path(dir, 'ICB_Nathanson/ICB_Nathanson_expr.tsv'), sep="\t") %>% as.data.frame()
dim(dat_exp_nathanson)

# Get the subset of data with both clinical and expression data.
dat_sample_nathanson = dat_sample_nathanson[rownames(dat_sample_nathanson) %in% colnames(dat_exp_nathanson), ]

dim(dat_sample_nathanson)
dat_sample_nathanson <- dat_sample_nathanson[order(rownames(dat_sample_nathanson)), ]
dat_exp_nathanson <- dat_exp_nathanson[, order(colnames(dat_exp_nathanson))]


####################################
### Snyder study
####################################

# clinicopathological characteristics  
dat_sample_snyder = read.table(file.path(dir, 'ICB_Snyder/ICB_Snyder_metadata.tsv'), sep="\t") %>% as.data.frame()
dim(dat_sample_snyder)

# gene meta data
dat_gene_snyder = read.table(file.path(dir, 'ICB_Snyder/ICB_Snyder_expr_genes.tsv'), sep="\t") %>% as.data.frame()
dim(dat_gene_snyder)

# RNA-seq expression data (i.e., log2(TPM + 0.001)) 
dat_exp_snyder = read.table(file.path(dir, 'ICB_Snyder/ICB_Snyder_expr.tsv'), sep="\t") %>% as.data.frame()
dim(dat_exp_snyder)

# Get the subset of data with both clinical and expression data.
dat_sample_snyder = dat_sample_snyder[rownames(dat_sample_snyder) %in% colnames(dat_exp_snyder), ]

dim(dat_sample_snyder)

dat_sample_snyder <- dat_sample_snyder[order(rownames(dat_sample_snyder)), ]
dat_exp_snyder <- dat_exp_snyder[, order(colnames(dat_exp_snyder))]

#######################################
### Van Allen study
#######################################

# clinicopathological characteristics  
dat_sample_van_allen = read.table(file.path(dir, 'ICB_Van_Allen/ICB_Van_Allen_metadata.tsv'), sep="\t") %>% as.data.frame()
dim(dat_sample_van_allen)

# gene meta data
dat_gene_van_allen = read.table(file.path(dir, 'ICB_Van_Allen/ICB_Van_Allen_expr_genes.tsv'), sep="\t") %>% as.data.frame()
dim(dat_gene_van_allen)

# RNA-seq expression data (i.e., log2(TPM + 0.001)) 
dat_exp_van_allen = read.table(file.path(dir, 'ICB_Van_Allen/ICB_Van_Allen_expr.tsv'), sep="\t") %>% as.data.frame()
dim(dat_exp_van_allen)

# Get the subset of data with both clinical and expression data.

dat_sample_van_allen = dat_sample_van_allen[rownames(dat_sample_van_allen) %in% colnames(dat_exp_van_allen), ]

dim(dat_sample_van_allen)

dat_sample_van_allen <- dat_sample_van_allen[order(rownames(dat_sample_van_allen)), ]
dat_exp_van_allen <- dat_exp_van_allen[, order(colnames(dat_exp_van_allen))]

################################
### Hugo study
################################

# clinicopathological characteristics  
dat_sample_hugo = read.table(file.path(dir, 'ICB_Hugo/ICB_Hugo_metadata.tsv'), sep="\t") %>% as.data.frame()
dim(dat_sample_hugo)

# gene meta data
dat_gene_hugo = read.table(file.path(dir, 'ICB_Hugo/ICB_Hugo_expr_gene_tpm_genes.tsv'), sep="\t") %>% as.data.frame()
dim(dat_gene_hugo)

# RNA-seq expression data (i.e., log2(TPM + 0.001)) 
dat_exp_hugo = read.table(file.path(dir, 'ICB_Hugo/ICB_Hugo_expr_gene_tpm.tsv'), sep="\t") %>% as.data.frame()
dim(dat_exp_hugo)

# Get the subset of data with both clinical and expression data.

dat_sample_hugo = dat_sample_hugo[rownames(dat_sample_hugo) %in% colnames(dat_exp_hugo), ]

dim(dat_sample_hugo)

dat_sample_hugo <- dat_sample_hugo[order(rownames(dat_sample_hugo)), ]
dat_exp_hugo <- dat_exp_hugo[, order(colnames(dat_exp_hugo))]


##########################
### Miao1 study
##########################

# clinicopathological characteristics  
dat_sample_miao1 = read.table(file.path(dir, 'ICB_Miao1/ICB_Miao1_metadata.tsv'), sep="\t") %>% as.data.frame()
dim(dat_sample_miao1)

# gene meta data
dat_gene_miao1 = read.table(file.path(dir, 'ICB_Miao1/ICB_Miao1_expr_genes.tsv'), sep="\t") %>% as.data.frame()
dim(dat_gene_miao1)

# RNA-seq expression data (i.e., log2(TPM + 0.001)) 
dat_exp_miao1 = read.table(file.path(dir, 'ICB_Miao1/ICB_Miao1_expr.tsv'), sep="\t") %>% as.data.frame()
dim(dat_exp_miao1)

# Get the subset of data with both clinical and expression data.

dat_sample_miao1 = dat_sample_miao1[rownames(dat_sample_miao1) %in% colnames(dat_exp_miao1), ]

dim(dat_sample_miao1)

dat_sample_miao1 <- dat_sample_miao1[order(rownames(dat_sample_miao1)), ]
dat_exp_miao1 <- dat_exp_miao1[, order(colnames(dat_exp_miao1))]

#########################
### Riaz study
#########################

# clinicopathological characteristics  
dat_sample_riaz = read.table(file.path(dir, 'ICB_Riaz/ICB_Riaz_metadata.tsv'), sep="\t") %>% as.data.frame()
dim(dat_sample_riaz)

# gene meta data
dat_gene_riaz = read.table(file.path(dir, 'ICB_Riaz/ICB_Riaz_expr_gene_tpm_genes.tsv'), sep="\t") %>% as.data.frame()
dim(dat_gene_riaz)

# RNA-seq expression data (i.e., log2(TPM + 0.001)) 
dat_exp_riaz = read.table(file.path(dir, 'ICB_Riaz/ICB_Riaz_expr_gene_tpm.tsv'), sep="\t") %>% as.data.frame()
dim(dat_exp_riaz)

# Get the subset of data with both clinical and expression data.

dat_sample_riaz = dat_sample_riaz[rownames(dat_sample_riaz) %in% colnames(dat_exp_riaz), ]

dim(dat_sample_riaz)

dat_sample_riaz <- dat_sample_riaz[order(rownames(dat_sample_riaz)), ]
dat_exp_riaz <- dat_exp_riaz[, order(colnames(dat_exp_riaz))]

#########################
### Gide study
#########################

# clinicopathological characteristics  
dat_sample_gide = read.table(file.path(dir, 'ICB_Gide/ICB_Gide_metadata.tsv'), sep="\t") %>% as.data.frame()
dim(dat_sample_gide)

# gene meta data
dat_gene_gide = read.table(file.path(dir, 'ICB_Gide/ICB_Gide_expr_gene_tpm_genes.tsv'), sep="\t") %>% as.data.frame()
dim(dat_gene_gide)

# RNA-seq expression data (i.e., log2(TPM + 0.001)) 
dat_exp_gide = read.table(file.path(dir, 'ICB_Gide/ICB_Gide_expr_gene_tpm.tsv'), sep="\t") %>% as.data.frame()
dim(dat_exp_gide)

# Get the subset of data with both clinical and expression data.

dat_sample_gide = dat_sample_gide[rownames(dat_sample_gide) %in% colnames(dat_exp_gide), ]

dim(dat_sample_gide)

dat_sample_gide <- dat_sample_gide[order(rownames(dat_sample_gide)), ]
dat_exp_gide <- dat_exp_gide[, order(colnames(dat_exp_gide))]

#########################
### Kim study
#########################

# clinicopathological characteristics  
dat_sample_kim = read.table(file.path(dir, 'ICB_kim/ICB_kim_metadata.tsv'), sep="\t") %>% as.data.frame()
dim(dat_sample_kim)

# gene meta data
dat_gene_kim = read.table(file.path(dir, 'ICB_kim/ICB_Kim_expr_gene_tpm_genes.tsv'), sep="\t") %>% as.data.frame()
dim(dat_gene_kim)

# RNA-seq expression data (i.e., log2(TPM + 0.001)) 
dat_exp_kim = read.table(file.path(dir, 'ICB_kim/ICB_Kim_expr_gene_tpm.tsv'), sep="\t") %>% as.data.frame()
dim(dat_exp_kim)

# Get the subset of data with both clinical and expression data.

dat_sample_kim = dat_sample_kim[rownames(dat_sample_kim) %in% colnames(dat_exp_kim), ]

dim(dat_sample_kim)

dat_sample_kim <- dat_sample_kim[order(rownames(dat_sample_kim)), ]
dat_exp_kim <- dat_exp_kim[, order(colnames(dat_exp_kim))]

##########################
### VanDenEnde study
##########################

# clinicopathological characteristics  
dat_sample_vanDenEnde = read.table(file.path(dir, 'ICB_vanDenEnde/ICB_vanDenEnde_metadata.tsv'), sep="\t") %>% as.data.frame()
dim(dat_sample_vanDenEnde)

# gene meta data
dat_gene_vanDenEnde = read.table(file.path(dir, 'ICB_vanDenEnde/ICB_vanDenEnde_expr_genes.tsv'), sep="\t") %>% as.data.frame()
dim(dat_gene_vanDenEnde)

# RNA-seq expression data (i.e., log2(TPM + 0.001)) 
dat_exp_vanDenEnde = read.table(file.path(dir, 'ICB_vanDenEnde/ICB_vanDenEnde_expr.tsv'), sep="\t") %>% as.data.frame()
dim(dat_exp_vanDenEnde)

# Get the subset of data with both clinical and expression data.

dat_sample_vanDenEnde = dat_sample_vanDenEnde[rownames(dat_sample_vanDenEnde) %in% colnames(dat_exp_vanDenEnde), ]

dim(dat_sample_vanDenEnde)

dat_sample_vanDenEnde <- dat_sample_vanDenEnde[order(rownames(dat_sample_vanDenEnde)), ]
dat_exp_vanDenEnde <- dat_exp_vanDenEnde[, order(colnames(dat_exp_vanDenEnde))]


#############################################################################################
## merge studies as list
#############################################################################################

dat_exp <- list(Braun = dat_exp_braun, 
                Hugo = dat_exp_hugo,
                Van_allen = dat_exp_van_allen,
                Miao1 = dat_exp_miao1,
                Nathanson = dat_exp_nathanson,
                Riaz = dat_exp_riaz,
                Snyder = dat_exp_snyder,
                Gide = dat_exp_gide,
                Kim = dat_exp_kim,
                VanDenEnde = dat_exp_vanDenEnde)


pdata <- list(Braun = dat_sample_braun, 
              Hugo = dat_sample_hugo,
              Van_allen = dat_sample_van_allen,
              Miao1 = dat_sample_miao1,
              Nathanson = dat_sample_nathanson,
              Riaz = dat_sample_riaz,
              Snyder = dat_sample_snyder,
              Gide = dat_sample_gide,
              Kim = dat_sample_kim,
              VanDenEnde = dat_sample_vanDenEnde)


```


## Logistic regression analysis 

For each study, we assess the association between TBC1D22B (ENSG00000065491.8) and ICB response (R vs NR) by fitting the logistic regression model. We fit the model under each study to identify the log(OR) and its standard error (SE). 

```{r logistic regression}

logit_res <- lapply(1:length(dat_exp), function(k){
  
  sub_dat_exp <- dat_exp[[k]]
  sub_pdata <- pdata[[k]]
  
  df <- data.frame(x = scale(as.numeric(sub_dat_exp[rownames(sub_dat_exp) == "ENSG00000065491.8",]))[,1],
                   y = sub_pdata$response)

  df <- df[!is.na(df$y),]
  df$y <- factor(df$y)
  df$y <- ifelse(df$y == "R", 0, 1)

  fit <- glm( y ~ x, 
            data = df, 
            family = "binomial")

fit_summary <- summary(fit)

 data.frame(gene_id = "ENSG00000065491.8",
           "logOR" = fit_summary$coefficients["x", "Estimate"],
           "SE" = fit_summary$coefficients["x", "Std. Error"],
           "pvals" = fit_summary$coefficients["x", "Pr(>|z|)"])
 })


logit_res <- do.call(rbind, logit_res)
rownames(logit_res) <- names(dat_exp)
  
```

## Meta-analysis

Meta-analyses can be performed to integrate independent studies to identify more reliable findings by increasing statistical power and reducing false positives. The strengths and limitations of meta-analyses are
evaluated particularly with respect to their ability to assess variation across studies or heterogeneity. Several meta-analysis methods have been proposed such as combining p-values, combining effect estimates, and rankings. 

Approaches that combine effects including fixed- and random-effects (FE & RE) models are widely used to achieve a broad inferential basis for evaluations of effects. Under the FE model, we assume that there is one true effect that underlies all the studies in the analysis, and that all differences in
observed effects are due to sampling error. In contrast, the RE model incorporates the variability of the effects across studies in addition to the within-study variability using a two-stage hierarchical process. 

Assessing heterogeneity is a critical issue in meta-analysis because different models may lead to different estimates of overall effect and different standard errors.Several approaches have been suggested over the years for how best to assess homogeneity across studies, and the corresponding research is ongoing. DerSimonian and Laird (DL) (or chi2) is widely used to assess the heterogeneity. 

### Part I: response (R vs. NR)

The meta-analysis is applied to integrate the results (i.e., logOR and SE) across ICB studies.
The forest plot demonstrates the fitted logistic regression model under each study including  the logOR, SE and p-value. 

The forest plot shows the estimate heterogeneity across studies $I^2 = 36\%$. In addition, both FE ($p = 0.21$) and RE ($p=0.65$) models show no association between TBC1D22B and ICB response.      

```{r meta analysis logistic resgression, fig.width = 7}

dat_meta <- logit_res

x <-  dat_meta$logOR
y <- dat_meta$SE
pvalue <- dat_meta$pvals

dat.xy <- data.frame(x, y, pvalue)
rownames(dat.xy) <- rownames(dat_meta)
res <- metagen(x, y, studlab = rownames(dat.xy), 
                    method.tau = "DL")


forest.meta(res,
            leftcols = c("studlab", "TE", "seTE", "pval"),
            leftlabs = c("ICB study/TBC1D22B", "logOR", "SE", "P value"),
            xlab = "logOR", 
            label.e = "Intervention",
            sortvar = TE, 
            #pooled.totals = TRUE,
            smlab = " ",
            text.random = "Random effect",
            print.I2.ci = FALSE,
            print.Q = TRUE,
            print.pval.Q = TRUE,
            digits.sd = 2,
            print.I2 = TRUE,
            print.tau2 = TRUE,
            comb.random =TRUE,
            comb.fixed = TRUE,
            text.fixed.w = TRUE,
            colgap.forest.left = "2cm",
            layout = "RevMan5", 
            test.overall.random = TRUE,
            test.overall.fixed = TRUE,
            xlim = "symmetric",
            col.square = "grey70",
            col.inside = "grey70",
            col.square.lines = "grey30",
            col.diamond.random = "blue2",
            col.diamond.fixed  = "red2",
            ff.xlab = "bold",
            fontsize = 11,
            fs.heading = 11.5,
            squaresize = 0.55,
            scientific.pval = TRUE,
            lty.random = NULL,
            lty.fixed  = NULL)


```

