---
title: "Signature-analysis: Immune-checkpoint blockade (ICB) data"
#author: "BHK lab"
date: '2022-09-07'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(meta)
library(metafor)
library(ggplot2)
library(survival)
library(survminer)
library(readr)
library(magrittr)
library(forestplot)
library(ggrepel)
library(GSVA)

```


## Load data


To download and extract TSV version of ICB data from Zenodo (https://zenodo.org/record/7058399/files/), follow the script named download_tsv_data R file at "~/Code/Download_Data".  

We read RNA-seq expression, clinicopathological characteristics, and gene meta data for the following selected ICB studies to include in meta-analysis. 

1. Braun (PMID 32472114)
2. Gide (PMID 30753825)
3. Hugo (PMID 26997480)
4. Miao1 (PMID 29301960)
5. Nathanson (PMID 27956380) 
6. Riaz (PMID 29301960)
7. Snyder (PMID 28552987) 
8. Van Allen (PMID 26359337)

As an example, we share the codes to get read the RNA-seq expression, clinicopathological characteristics, and gene meta data usign ICB Braun data. The Braun study includes 319 patients with clinical data and 40,994 genes. While only 181 patients have both clinical data and RNA-seq expression data. 

Note that the expression data is $log2(\text{TPM} + 0.001)$, where TPM represents transcripts per million. 

### Braun study

```{r load ICB data Braun}

# clinicopathological characteristics  
dat_sample_braun = read.table('~/ICB_Braun/ICB_Braun_metadata.tsv', sep="\t") %>% as.data.frame()
dim(dat_sample_braun)

# gene meta data
dat_gene_braun = read.table('~/ICB_Braun/ICB_Braun_expr_genes.tsv', sep="\t") %>% as.data.frame()
dim(dat_gene_braun)

# RNA-seq expression data (i.e., log2(TPM + 0.001)) 
dat_exp_braun = read.table('~/ICB_Braun/ICB_Braun_expr.tsv', sep="\t") %>% as.data.frame()
dim(dat_exp_braun)

# Get the subset of data with both clinical and expression data.

dat_sample_braun = dat_sample_braun[rownames(dat_sample_braun) %in% colnames(dat_exp_braun), ]

dim(dat_sample_braun)

dat_sample_braun <- dat_sample_braun[order(rownames(dat_sample_braun)), ]
dat_exp_braun <- dat_exp_braun[, order(colnames(dat_exp_braun))]


dat_gene_braun <- dat_gene_braun[order(dat_gene_braun$gene_id), ]
dat_gene_braun <- dat_gene_braun[!duplicated(dat_gene_braun$gene_name), ]
dat_exp_braun <- dat_exp_braun[rownames(dat_exp_braun) %in% dat_gene_braun$gene_id, ]

```

```{r load ICB data, include=FALSE}

##################################
### Nathanson study
##################################
# clinicopathological characteristics  
dat_sample_nathanson = read.table('~/ICB_Nathanson/ICB_Nathanson_metadata.tsv', sep="\t") %>% as.data.frame()
dim(dat_sample_nathanson)

# gene meta data
dat_gene_nathanson = read.table('~/ICB_Nathanson/ICB_Nathanson_expr_genes.tsv', sep="\t") %>% as.data.frame()
dim(dat_gene_nathanson)

# RNA-seq expression data (i.e., log2(TPM + 0.001)) 
dat_exp_nathanson = read.table('~/ICB_Nathanson/ICB_Nathanson_expr.tsv', sep="\t") %>% as.data.frame()
dim(dat_exp_nathanson)

# Get the subset of data with both clinical and expression data.
dat_sample_nathanson = dat_sample_nathanson[rownames(dat_sample_nathanson) %in% colnames(dat_exp_nathanson), ]

dim(dat_sample_nathanson)
dat_sample_nathanson <- dat_sample_nathanson[order(rownames(dat_sample_nathanson)), ]
dat_exp_nathanson <- dat_exp_nathanson[, order(colnames(dat_exp_nathanson))]

dat_gene_nathanson <- dat_gene_nathanson[order(dat_gene_nathanson$gene_id), ]
dat_gene_nathanson <- dat_gene_nathanson[!duplicated(dat_gene_nathanson$gene_name), ]
dat_exp_nathanson <- dat_exp_nathanson[rownames(dat_exp_nathanson) %in% dat_gene_nathanson$gene_id, ]

####################################
### Snyder study
####################################

# clinicopathological characteristics  
dat_sample_snyder = read.table('~/ICB_Snyder/ICB_Snyder_metadata.tsv', sep="\t") %>% as.data.frame()
dim(dat_sample_snyder)

# gene meta data
dat_gene_snyder = read.table('~/ICB_Snyder/ICB_Snyder_expr_genes.tsv', sep="\t") %>% as.data.frame()
dim(dat_gene_snyder)

# RNA-seq expression data (i.e., log2(TPM + 0.001)) 
dat_exp_snyder = read.table('~/ICB_Snyder/ICB_Snyder_expr.tsv', sep="\t") %>% as.data.frame()
dim(dat_exp_snyder)

# Get the subset of data with both clinical and expression data.
dat_sample_snyder = dat_sample_snyder[rownames(dat_sample_snyder) %in% colnames(dat_exp_snyder), ]

dim(dat_sample_snyder)

dat_sample_snyder <- dat_sample_snyder[order(rownames(dat_sample_snyder)), ]
dat_exp_snyder <- dat_exp_snyder[, order(colnames(dat_exp_snyder))]

dat_gene_snyder <- dat_gene_snyder[order(dat_gene_snyder$gene_id), ]
dat_gene_snyder <- dat_gene_snyder[!duplicated(dat_gene_snyder$gene_name), ]
dat_exp_snyder <- dat_exp_snyder[rownames(dat_exp_snyder) %in% dat_gene_snyder$gene_id, ]

#######################################
### Van Allen study
#######################################

# clinicopathological characteristics  
dat_sample_van_allen = read.table('~/ICB_Van_Allen/ICB_Van_Allen_metadata.tsv', sep="\t") %>% as.data.frame()
dim(dat_sample_van_allen)

# gene meta data
dat_gene_van_allen = read.table('~/ICB_Van_Allen/ICB_Van_Allen_expr_genes.tsv', sep="\t") %>% as.data.frame()
dim(dat_gene_van_allen)

# RNA-seq expression data (i.e., log2(TPM + 0.001)) 
dat_exp_van_allen = read.table('~/ICB_Van_Allen/ICB_Van_Allen_expr.tsv', sep="\t") %>% as.data.frame()
dim(dat_exp_van_allen)

# Get the subset of data with both clinical and expression data.

dat_sample_van_allen = dat_sample_van_allen[rownames(dat_sample_van_allen) %in% colnames(dat_exp_van_allen), ]

dim(dat_sample_van_allen)

dat_sample_van_allen <- dat_sample_van_allen[order(rownames(dat_sample_van_allen)), ]
dat_exp_van_allen <- dat_exp_van_allen[, order(colnames(dat_exp_van_allen))]

dat_gene_van_allen <- dat_gene_van_allen[order(dat_gene_van_allen$gene_id), ]
dat_gene_van_allen <- dat_gene_van_allen[!duplicated(dat_gene_van_allen$gene_name), ]
dat_exp_van_allen <- dat_exp_van_allen[rownames(dat_exp_van_allen) %in% dat_gene_van_allen$gene_id, ]

################################
### Hugo study
################################

# clinicopathological characteristics  
dat_sample_hugo = read.table('~/ICB_Hugo/ICB_Hugo_metadata.tsv', sep="\t") %>% as.data.frame()
dim(dat_sample_hugo)

# gene meta data
dat_gene_hugo = read.table('~/ICB_Hugo/ICB_Hugo_expr_gene_tpm_genes.tsv', sep="\t") %>% as.data.frame()
dim(dat_gene_hugo)

# RNA-seq expression data (i.e., log2(TPM + 0.001)) 
dat_exp_hugo = read.table('~/ICB_Hugo/ICB_Hugo_expr_gene_tpm.tsv', sep="\t") %>% as.data.frame()
dim(dat_exp_hugo)

# Get the subset of data with both clinical and expression data.

dat_sample_hugo = dat_sample_hugo[rownames(dat_sample_hugo) %in% colnames(dat_exp_hugo), ]

dim(dat_sample_hugo)

dat_sample_hugo <- dat_sample_hugo[order(rownames(dat_sample_hugo)), ]
dat_exp_hugo <- dat_exp_hugo[, order(colnames(dat_exp_hugo))]


dat_gene_hugo <- dat_gene_hugo[order(dat_gene_hugo$gene_id), ]
dat_gene_hugo <- dat_gene_hugo[!duplicated(dat_gene_hugo$gene_name), ]
dat_exp_hugo <- dat_exp_hugo[rownames(dat_exp_hugo) %in% dat_gene_hugo$gene_id, ]

##########################
### Miao1 study
##########################

# clinicopathological characteristics  
dat_sample_miao1 = read.table('~/ICB_Miao1/ICB_Miao1_metadata.tsv', sep="\t") %>% as.data.frame()
dim(dat_sample_miao1)

# gene meta data
dat_gene_miao1 = read.table('~/ICB_Miao1/ICB_Miao1_expr_genes.tsv', sep="\t") %>% as.data.frame()
dim(dat_gene_miao1)

# RNA-seq expression data (i.e., log2(TPM + 0.001)) 
dat_exp_miao1 = read.table('~/ICB_Miao1/ICB_Miao1_expr.tsv', sep="\t") %>% as.data.frame()
dim(dat_exp_miao1)

# Get the subset of data with both clinical and expression data.

dat_sample_miao1 = dat_sample_miao1[rownames(dat_sample_miao1) %in% colnames(dat_exp_miao1), ]

dim(dat_sample_miao1)

dat_sample_miao1 <- dat_sample_miao1[order(rownames(dat_sample_miao1)), ]
dat_exp_miao1 <- dat_exp_miao1[, order(colnames(dat_exp_miao1))]

dat_gene_miao1 <- dat_gene_miao1[order(dat_gene_miao1$gene_id), ]
dat_gene_miao1 <- dat_gene_miao1[!duplicated(dat_gene_miao1$gene_name), ]
dat_exp_miao1 <- dat_exp_miao1[rownames(dat_exp_miao1) %in% dat_gene_miao1$gene_id, ]

#########################
### Riaz study
#########################

# clinicopathological characteristics  
dat_sample_riaz = read.table('~/ICB_Riaz/ICB_Riaz_metadata.tsv', sep="\t") %>% as.data.frame()
dim(dat_sample_riaz)

# gene meta data
dat_gene_riaz = read.table('~/ICB_Riaz/ICB_Riaz_expr_gene_tpm_genes.tsv', sep="\t") %>% as.data.frame()
dim(dat_gene_riaz)

# RNA-seq expression data (i.e., log2(TPM + 0.001)) 
dat_exp_riaz = read.table('~/ICB_Riaz/ICB_Riaz_expr_gene_tpm.tsv', sep="\t") %>% as.data.frame()
dim(dat_exp_riaz)

# Get the subset of data with both clinical and expression data.

dat_sample_riaz = dat_sample_riaz[rownames(dat_sample_riaz) %in% colnames(dat_exp_riaz), ]

dim(dat_sample_riaz)

dat_sample_riaz <- dat_sample_riaz[order(rownames(dat_sample_riaz)), ]
dat_exp_riaz <- dat_exp_riaz[, order(colnames(dat_exp_riaz))]

dat_gene_riaz <- dat_gene_riaz[order(dat_gene_riaz$gene_id), ]
dat_gene_riaz <- dat_gene_riaz[!duplicated(dat_gene_riaz$gene_name), ]
dat_exp_riaz <- dat_exp_riaz[rownames(dat_exp_riaz) %in% dat_gene_riaz$gene_id, ]

#########################
### Gide study
#########################

# clinicopathological characteristics  
dat_sample_gide = read.table('~/ICB_Gide/ICB_Gide_metadata.tsv', sep="\t") %>% as.data.frame()
dim(dat_sample_gide)

# gene meta data
dat_gene_gide = read.table('~/ICB_Gide/ICB_Gide_expr_gene_tpm_genes.tsv', sep="\t") %>% as.data.frame()
dim(dat_gene_gide)

# RNA-seq expression data (i.e., log2(TPM + 0.001)) 
dat_exp_gide = read.table('~/ICB_Gide/ICB_Gide_expr_gene_tpm.tsv', sep="\t") %>% as.data.frame()
dim(dat_exp_gide)

# Get the subset of data with both clinical and expression data.

dat_sample_gide = dat_sample_gide[rownames(dat_sample_gide) %in% colnames(dat_exp_gide), ]

dim(dat_sample_gide)

dat_sample_gide <- dat_sample_gide[order(rownames(dat_sample_gide)), ]
dat_exp_gide <- dat_exp_gide[, order(colnames(dat_exp_gide))]

dat_gene_gide <- dat_gene_gide[order(dat_gene_gide$gene_id), ]
dat_gene_gide <- dat_gene_gide[!duplicated(dat_gene_gide$gene_name), ]
dat_exp_gide <- dat_exp_gide[rownames(dat_exp_gide) %in% dat_gene_gide$gene_id, ]

#############################################################################################
## combine all studies as list
#############################################################################################

dat_exp <- list(Braun = dat_exp_braun, 
                Hugo = dat_exp_hugo,
                Van_allen = dat_exp_van_allen,
                Miao1 = dat_exp_miao1,
                Nathanson = dat_exp_nathanson,
                Riaz = dat_exp_riaz,
                Snyder = dat_exp_snyder,
                Gide = dat_exp_gide)

pdata <- list(Braun = dat_sample_braun, 
              Hugo = dat_sample_hugo,
              Van_allen = dat_sample_van_allen,
              Miao1 = dat_sample_miao1,
              Nathanson = dat_sample_nathanson,
              Riaz = dat_sample_riaz,
              Snyder = dat_sample_snyder,
              Gide = dat_sample_gide)


dat_gene <- list(Braun = dat_gene_braun, 
                Hugo = dat_gene_hugo,
                Van_allen = dat_gene_van_allen,
                Miao1 = dat_gene_miao1,
                Nathanson = dat_gene_nathanson,
                Riaz = dat_gene_riaz,
                Snyder = dat_gene_snyder,
                Gide = dat_gene_gide)



for(i in 1:length(dat_exp)){
  
  rownames(dat_exp[[i]]) <- dat_gene[[i]]$gene_name
  
}


```

## Signature data

To download signatures, consider the "~/Data/Signature" folder at GitHub repo. We make a list of available signatures. 

Under each study, we consider gene set variation analysis (GSVA) approach to compute the signature score. Then, we assess the association between signature score and overall survival (OS) using the Cox model. The meta-analysis is applied to integrate results (logHR and SE) across studies. The heterogeneity across studies is evaluated using the DL method. 


```{r signature analysis, warning=FALSE, message=FALSE}

signature_names <- c("CD8_Jiang", "T_cell_inflamed", "Inflammatory", "CYT",
                     "STAT1", "IRG_Ayers", "TLS", "IFNG", "PDL1")

dir0 <- "C:/ICBRoche/Signature/"
dir <- paste(dir0, paste(signature_names, "csv", sep="."), sep="")

geneSets <- lapply(1:length(dir), function(k){
  
  dat <- read.csv(dir[k])
  dat[,1]
  
})

names(geneSets) <- signature_names

## run the GSVA method to compute signature score 

gsva_res <- lapply(1:length(dat_exp), function(k){
  
 gsva(as.matrix(dat_exp[[k]]), geneSets, method = "gsva", verbose = FALSE)
 
})

names(gsva_res) <- names(dat_exp)

```


## Survival analysis (OS) 

For each selected ICB studies, we assess the association of computed signature score and overall survival (OS) by fitting the Cox model per study. Then, the meta-analysis is applied to integrate the results (i.e., logHR and SE) across ICB studies. 

```{r cox all genes, warning=FALSE, message=FALSE, include=FALSE}

cox_res <- lapply(1:length(gsva_res), function(k){
  
  sub_dat_exp <- gsva_res[[k]]
  sub_pdata <- pdata[[k]]
 
  res <- lapply(1:nrow(sub_dat_exp), function(j){
    
    df <- data.frame(x = as.numeric(sub_dat_exp[j,]),
                     t.os = sub_pdata$survival_time_os,
                     os = sub_pdata$event_occurred_os)
    
    fit <- coxph(Surv(t.os, os) ~ x, data = df)
    fit_summary <- summary(fit)
    
    data.frame(study = names(gsva_res)[k],
               signature_name = rownames(sub_dat_exp)[j],
               "logHR" = fit_summary$coefficients["x", "coef"],
               "SE" = fit_summary$coefficients["x", "se(coef)"],
               "pvals" = fit_summary$coefficients["x", "Pr(>|z|)"])
    
  })
  
  do.call(rbind, res)

})

cox_res <- do.call(rbind, cox_res)

  
```


## Meta-analysis

Meta-analyses can be performed to integrate independent studies to identify more reliable findings by increasing statistical power and reducing false positives. The strengths and limitations of meta-analyses are
evaluated particularly with respect to their ability to assess variation across studies or heterogeneity. Several meta-analysis methods have been proposed such as combining p-values, combining effect estimates, and rankings. 

Approaches that combine effects including fixed- and random-effects (FE & RE) models are widely used to achieve a broad inferential basis for evaluations of effects. Under the FE model, we assume that there is one true effect that underlies all the studies in the analysis, and that all differences in
observed effects are due to sampling error. In contrast, the RE model incorporates the variability of the effects across studies in addition to the within-study variability using a two-stage hierarchical process. 

Assessing heterogeneity is a critical issue in meta-analysis because different models may lead to different estimates of overall effect and different standard errors.Several approaches have been suggested over the years for how best to assess homogeneity across studies, and the corresponding research is ongoing. DerSimonian and Laird (DL) (or chi2) is widely used to assess the heterogeneity. 

### Part I: Overall survival (OS)

The meta-analysis was applied to integrate the results (logHR, SE) across studies using DL method to estimate heterogeneity. Signature CYT (logHR = -0.38, p = 0.04) and PDL1 (logHR = -0.42, p = 0.03) are associated with OS. In addition, for the associated signatures with OS, the forestplots show significant heterogeneity across studies as well: CYT ($I^2 = 62\%$ and $p = 9.5e-03$) and PDL1 ($I^2 = 59\%$ and $p = 1.8e-02$).  

```{r meta analysis logistic resgression, fig.width = 7, warning=FALSE, message=FALSE, echo=FALSE}

meta_res <- lapply(1:length(signature_names), function(k){

  dat_meta <- cox_res[cox_res$signature_name == signature_names[k], ]
  x <-  dat_meta$logHR
  y <- dat_meta$SE
  pvalue <- dat_meta$pvals

  dat.xy <- data.frame(x, y, pvalue)
  rownames(dat.xy) <- rownames(dat_meta)
  res <- metagen(x, y, studlab = rownames(dat.xy), 
                    method.tau = "DL")

 data.frame(signature_name = dat_meta$signature_name[1],
            TE_RE = res$TE.random,
            pval_RE = res$pval.random,
            I2 = res$I2,
            Q_pval = res$pval.Q)

})

meta_res  <- do.call(rbind, meta_res )

meta_res

```


```{r forestplot CYT signature, echo =FALSE}

dat_meta <- cox_res[cox_res$signature_name == "CYT", ]
  x <-  dat_meta$logHR
  y <- dat_meta$SE
  pvalue <- dat_meta$pvals

  dat.xy <- data.frame(x, y, pvalue)
  rownames(dat.xy) <- dat_meta$study
  res <- metagen(x, y, studlab = rownames(dat.xy), 
                    method.tau = "DL")


forest.meta(res,
            leftcols = c("studlab", "TE", "seTE", "pval"),
            leftlabs = c("ICB study/CYT", "logOR", "SE", "P value"),
            xlab = "logOR", 
            label.e = "Intervention",
            sortvar = TE, 
            #pooled.totals = TRUE,
            smlab = " ",
            text.random = "Random effect",
            print.I2.ci = FALSE,
            print.Q = TRUE,
            print.pval.Q = TRUE,
            digits.sd = 2,
            print.I2 = TRUE,
            print.tau2 = TRUE,
            comb.random =TRUE,
            comb.fixed = TRUE,
            text.fixed.w = TRUE,
            colgap.forest.left = "2cm",
            layout = "RevMan5", 
            test.overall.random = TRUE,
            test.overall.fixed = TRUE,
            xlim = "symmetric",
            col.square = "grey70",
            col.inside = "grey70",
            col.square.lines = "grey30",
            col.diamond.random = "blue2",
            col.diamond.fixed  = "red2",
            ff.xlab = "bold",
            fontsize = 11,
            fs.heading = 11.5,
            squaresize = 0.55,
            scientific.pval = TRUE,
            lty.random = NULL,
            lty.fixed  = NULL)

```



```{r forestplot PDL1 signature, echo =FALSE}

dat_meta <- cox_res[cox_res$signature_name == "PDL1", ]
  x <-  dat_meta$logHR
  y <- dat_meta$SE
  pvalue <- dat_meta$pvals

  dat.xy <- data.frame(x, y, pvalue)
  rownames(dat.xy) <- dat_meta$study
  res <- metagen(x, y, studlab = rownames(dat.xy), 
                    method.tau = "DL")


forest.meta(res,
            leftcols = c("studlab", "TE", "seTE", "pval"),
            leftlabs = c("ICB study/PDL1", "logOR", "SE", "P value"),
            xlab = "logOR", 
            label.e = "Intervention",
            sortvar = TE, 
            #pooled.totals = TRUE,
            smlab = " ",
            text.random = "Random effect",
            print.I2.ci = FALSE,
            print.Q = TRUE,
            print.pval.Q = TRUE,
            digits.sd = 2,
            print.I2 = TRUE,
            print.tau2 = TRUE,
            comb.random =TRUE,
            comb.fixed = TRUE,
            text.fixed.w = TRUE,
            colgap.forest.left = "2cm",
            layout = "RevMan5", 
            test.overall.random = TRUE,
            test.overall.fixed = TRUE,
            xlim = "symmetric",
            col.square = "grey70",
            col.inside = "grey70",
            col.square.lines = "grey30",
            col.diamond.random = "blue2",
            col.diamond.fixed  = "red2",
            ff.xlab = "bold",
            fontsize = 11,
            fs.heading = 11.5,
            squaresize = 0.55,
            scientific.pval = TRUE,
            lty.random = NULL,
            lty.fixed  = NULL)

```

